---
title: "Tugas Pertemuan 3"
author: "Nabila Yola Zahrani M0401241045"
date: "2026-09-03"
output: html_document
---

# [No 1 - Distributed Lag]{style="color:#db84c5;"} 
## [Data]{style="color:#db84c5;"}
```{r}
data1 <- data.frame(
  periode = 1:20,
  market_share = c(
    3.63, 4.20, 3.33, 4.54, 2.89,
    4.87, 4.90, 5.29, 6.18, 7.20,
    7.25, 6.09, 6.80, 8.65, 8.43,
    8.29, 7.18, 7.90, 8.45, 8.23
  ),
  price = c(
    0.97, 0.95, 0.99, 0.91, 0.98,
    0.90, 0.89, 0.86, 0.85, 0.82,
    0.79, 0.83, 0.81, 0.77, 0.76,
    0.80, 0.83, 0.79, 0.76, 0.78
  )
)

data1
```

## [Membuat Peubah Lag]{style="color:#db84c5;"}
```{r}
data1$price_lag1 <- c(NA, data1$price[-20])
data1$price_lag2 <- c(NA, NA, data1$price[1:18])
data1$price_lag3 <- c(NA, NA, NA, data1$price[1:17])
data1$price_lag4 <- c(NA, NA, NA, NA, data1$price[1:16])

data1
```

## [Model Distributed Lag 0]{style="color:#db84c5;"}
```{r}
model_DL0 <- lm(
  market_share ~ price,
  data = data1
)

summary(model_DL0)
```

## [Model Distributed Lag 1]{style="color:#db84c5;"}
```{r}
model_DL1 <- lm(
  market_share ~ price + price_lag1,
  data = data1
)

summary(model_DL1)
```

## [Model Distributed Lag 2]{style="color:#db84c5;"}
```{r}
model_DL2 <- lm(
  market_share ~ price + price_lag1 + price_lag2,
  data = data1
)

summary(model_DL2)
```

## [Model Distributed Lag 3]{style="color:#db84c5;"}
```{r}
model_DL3 <- lm(
  market_share ~ price + price_lag1 + price_lag2 + price_lag3,
  data = data1
)

summary(model_DL3)
```

## [Model Distributed Lag 4]{style="color:#db84c5;"}
```{r}
model_DL4 <- lm(
  market_share ~ price + price_lag1 + price_lag2 + price_lag3 + price_lag4,
  data = data1
)

summary(model_DL4)
```


## [Evaluasi Model]{style="color:#db84c5;"}
```{r}
RMSE_DL0 <- sqrt(mean(residuals(model_DL0)^2))
RMSE_DL1 <- sqrt(mean(residuals(model_DL1)^2))
RMSE_DL2 <- sqrt(mean(residuals(model_DL2)^2))
RMSE_DL3 <- sqrt(mean(residuals(model_DL3)^2))
RMSE_DL4 <- sqrt(mean(residuals(model_DL4)^2))

MAE_DL0 <- mean(abs(residuals(model_DL0)))
MAE_DL1 <- mean(abs(residuals(model_DL1)))
MAE_DL2 <- mean(abs(residuals(model_DL2)))
MAE_DL3 <- mean(abs(residuals(model_DL3)))
MAE_DL4 <- mean(abs(residuals(model_DL4)))

hasil_DL <- data.frame(
  Model = c("DL(0)", "DL(1)", "DL(2)", "DL(3)", "DL(4)"),
  RMSE = c(
    RMSE_DL0,
    RMSE_DL1,
    RMSE_DL2,
    RMSE_DL3,
    RMSE_DL4
  ),
  MAE = c(
    MAE_DL0,
    MAE_DL1,
    MAE_DL2,
    MAE_DL3,
    MAE_DL4
  ),
  AIC = c(
    AIC(model_DL0),
    AIC(model_DL1),
    AIC(model_DL2),
    AIC(model_DL3),
    AIC(model_DL4)
  )
)

hasil_DL
```

## [Model Terbaik]{style="color:#db84c5;"}
```{r}
hasil_DL[which.min(hasil_DL$RMSE), ]
```


# [No 2 - ARDL]{style="color:#db84c5;"} 

## [Memasukan Data]{style="color:#db84c5;"}
```{r}
data2 <- data.frame(
  year = 1975:2014,
  sales = c(
    11.7, 12.0, 12.3, 12.8, 13.1,
    13.6, 13.9, 14.4, 14.7, 15.3,
    15.5, 15.8, 16.1, 16.6, 16.9,
    16.7, 16.9, 17.4, 17.6, 17.9,
    18.0, 17.9, 18.0, 18.2, 18.2,
    18.3, 18.6, 19.2, 19.3, 19.5,
    19.2, 19.3, 19.5, 20.0, 20.0,
    19.9, 19.8, 19.9, 20.2, 21.0
  ),
  advertising = c(
    9.4, 9.6, 10.0, 10.4, 10.8,
    10.9, 11.7, 12.2, 12.5, 12.9,
    13.0, 13.2, 13.8, 14.2, 14.6,
    14.4, 15.0, 15.4, 15.7, 15.9,
    15.9, 16.0, 16.3, 16.2, 16.8,
    17.3, 17.6, 18.1, 18.3, 18.5,
    18.7, 18.9, 19.2, 20.0, 20.0,
    20.3, 20.4, 21.0, 21.5, 22.1
  )
)

data2
```

## [Membuat Peubah Lag]{style="color:#db84c5;"}
```{r}
data2$sales_lag1 <- c(NA, data2$sales[-40])
data2$sales_lag2 <- c(NA, NA, data2$sales[1:38])
data2$sales_lag3 <- c(NA, NA, NA, data2$sales[1:37])
data2$sales_lag4 <- c(NA, NA, NA, NA, data2$sales[1:36])

data2$advertising_lag1 <- c(NA, data2$advertising[-40])
data2$advertising_lag2 <- c(NA, NA, data2$advertising[1:38])
data2$advertising_lag3 <- c(NA, NA, NA, data2$advertising[1:37])
data2$advertising_lag4 <- c(NA, NA, NA, NA, data2$advertising[1:36])

data2
```


## [Model ARDL 1,1]{style="color:#db84c5;"}
```{r}
model_ARDL11 <- lm(
  sales ~ sales_lag1 + advertising + advertising_lag1,
  data = data2
)

summary(model_ARDL11)
```

## [Model ARDL 1,2]{style="color:#db84c5;"}
```{r}
model_ARDL12 <- lm(
  sales ~ sales_lag1 + advertising +
    advertising_lag1 + advertising_lag2,
  data = data2
)

summary(model_ARDL12)
```

## [Model ARDL 2,2]{style="color:#db84c5;"}
```{r}
model_ARDL22 <- lm(
  sales ~ sales_lag1 + sales_lag2 +
    advertising + advertising_lag1 + advertising_lag2,
  data = data2
)

summary(model_ARDL22)
```

## [Model ARDL 3,3]{style="color:#db84c5;"}
```{r}
model_ARDL33 <- lm(
  sales ~ sales_lag1 + sales_lag2 + sales_lag3 +
    advertising + advertising_lag1 +
    advertising_lag2 + advertising_lag3,
  data = data2
)

summary(model_ARDL33)
```

## [Model ARDL 4,4]{style="color:#db84c5;"}
```{r}
model_ARDL44 <- lm(
  sales ~ sales_lag1 + sales_lag2 + sales_lag3 + sales_lag4 +
    advertising + advertising_lag1 +
    advertising_lag2 + advertising_lag3 + advertising_lag4,
  data = data2
)

summary(model_ARDL44)
```

## [Evaluasi Model ARDL]{style="color:#db84c5;"}
```{r}
RMSE_11 <- sqrt(mean(residuals(model_ARDL11)^2))
RMSE_12 <- sqrt(mean(residuals(model_ARDL12)^2))
RMSE_22 <- sqrt(mean(residuals(model_ARDL22)^2))
RMSE_33 <- sqrt(mean(residuals(model_ARDL33)^2))
RMSE_44 <- sqrt(mean(residuals(model_ARDL44)^2))

MAE_11 <- mean(abs(residuals(model_ARDL11)))
MAE_12 <- mean(abs(residuals(model_ARDL12)))
MAE_22 <- mean(abs(residuals(model_ARDL22)))
MAE_33 <- mean(abs(residuals(model_ARDL33)))
MAE_44 <- mean(abs(residuals(model_ARDL44)))

hasil_ARDL <- data.frame(
  Model = c(
    "ARDL(1,1)",
    "ARDL(1,2)",
    "ARDL(2,2)",
    "ARDL(3,3)",
    "ARDL(4,4)"
  ),
  RMSE = c(
    RMSE_11,
    RMSE_12,
    RMSE_22,
    RMSE_33,
    RMSE_44
  ),
  MAE = c(
    MAE_11,
    MAE_12,
    MAE_22,
    MAE_33,
    MAE_44
  ),
  AIC = c(
    AIC(model_ARDL11),
    AIC(model_ARDL12),
    AIC(model_ARDL22),
    AIC(model_ARDL33),
    AIC(model_ARDL44)
  )
)

hasil_ARDL
```

## [Model Terbaik]{style="color:#db84c5;"}
```{r}
hasil_ARDL[which.min(hasil_ARDL$RMSE), ]
```


